---
title: "BC Gov Referral Spatial Analysis Report"
author: "BC Government Referral Process"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: false
    code-fold: true
    code-summary: "Show code"
params:
  geomark_input: "https://apps.gov.bc.ca/pub/geomark/geomarks/gm-723966ECBECD4FE698C92C9D079F2474"
  config_path: "referral_layers.xlsx"
  interactive_maps: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)

utils_files <- c(
  "R/utils_core.R",
  "R/utils_layers.R",
  "R/utils_analysis_polygon.R",
  "R/utils_analysis_line.R",
  "R/utils_analysis_proximity.R",
  "R/utils_analysis_special.R",
  "R/utils_report_helpers.R"
)
invisible(lapply(utils_files, source))
setup_packages()
```

```{html}
<style>
details {
  margin: 1.5em 0;
  padding: 0.5em 0.5em 0;
  border: 1px solid #ddd;
  border-radius: 4px;
  background-color: #f9f9f9;
}

details summary {
  cursor: pointer;
  font-weight: 600;
  color: #0066cc;
  padding: 0.5em;
  margin: -0.5em -0.5em 0 -0.5em;
  border-radius: 4px 4px 0 0;
  background-color: #f0f7ff;
  user-select: none;
}

details summary:hover {
  background-color: #e6f0ff;
}

details > *:not(summary) {
  margin: 0.5em 0;
}
</style>
<script>
document.addEventListener("DOMContentLoaded", function () {
  document.querySelectorAll("p").forEach(function (p) {
    if (p.textContent && p.textContent.trim() === "NULL") {
      p.remove();
    }
  });
});
</script>
```

# Overview

This report provides an automated spatial analysis of reference layers against
a referral Area of Interest (AOI). The AOI is loaded from a BC Geomark URL or
a local spatial file, and each configured reference layer is checked for
overlap or proximity.

- **Analysis date:** `r Sys.Date()`
- **Proximity analysis distance:** layer-specific values from `Distance_outside_plot` in the configured layers workbook
- **Map mode:** `r if (isTRUE(params$interactive_maps)) "Interactive (leaflet)" else "Static (ggplot2)"`

# Load AOI

```{r load-aoi}
if (params$geomark_input == "") {
  stop(
    "No geomark input provided. Set the 'geomark_input' parameter to a ",
    "geomark URL or local file path.\n\n",
    "Example usage:\n",
    "  quarto render referral_analysis.qmd ",
    "-P geomark_input:https://apps.gov.bc.ca/pub/geomark/geomarks/gm-abc123"
  )
}

aoi <- load_geomark(params$geomark_input)
aoi_geom_type <- get_geometry_type(aoi)
aoi_area_ha <- if (aoi_geom_type == "polygon") {
  as.numeric(sf::st_area(sf::st_union(aoi))) / 10000
} else {
  NA
}
aoi_length_km <- if (aoi_geom_type == "line") {
  as.numeric(sf::st_length(sf::st_union(aoi))) / 1000
} else {
  NA
}
```

**AOI geometry type:** `r aoi_geom_type`

```{r aoi-summary, results='asis'}
if (aoi_geom_type == "polygon") {
  cat(sprintf("**AOI area:** %.2f ha (%.4f km²)\n", aoi_area_ha, aoi_area_ha / 100))
} else if (aoi_geom_type == "line") {
  cat(sprintf("**AOI length:** %.2f km\n", aoi_length_km))
}
```

## AOI Map

```{r aoi-map, out.height="500px"}
if (isTRUE(params$interactive_maps)) {
  render_leaflet_widget(build_aoi_leaflet(aoi))
} else {
  print(
    ggplot2::ggplot() +
      ggplot2::geom_sf(data = aoi, fill = "steelblue", alpha = 0.3, colour = "navy") +
      ggplot2::theme_minimal() +
      ggplot2::labs(title = "Referral Area of Interest (AOI)")
  )
}
```

# Reference Layer Analysis

```{r load-config, results='asis'}
init_layers <- initialize_referral_layers(config_path = params$config_path)
config <- init_layers$config
cat(sprintf("**Configured layers:** %d\n\n", nrow(config)))
```

```{r run-analysis, results='asis'}
layer_outputs <- list()
results <- list()
no_hit_layers <- list()
skipped_layers <- list()

for (i in seq_len(nrow(config))) {
  layer_info <- config[i, ]
  layer_name <- as.character(layer_info$layer_name)
  layer_display <- if ("name" %in% names(layer_info)) as.character(layer_info$name) else layer_name
  layer_path <- as.character(layer_info$layer_path)
  layer_geom <- as.character(layer_info$geometry_type)
  layer_focus <- if ("focus" %in% names(layer_info)) as.character(layer_info$focus) else "Unspecified"
  spatial_need <- if (aoi_geom_type == "polygon") {
    as.character(layer_info$spatial_information_polygon)
  } else {
    as.character(layer_info$spatial_information_linestring)
  }

  fields <- NULL
  if ("fields_to_extract" %in% names(layer_info)) {
    fields <- layer_info$fields_to_extract[[1]]
  }
  if (is.null(fields)) {
    fields <- parse_key_values(if ("key_values" %in% names(layer_info)) as.character(layer_info$key_values) else NA_character_)
  }

  outside_distance <- NA_real_
  if ("distance_outside_m" %in% names(layer_info)) {
    outside_distance <- suppressWarnings(as.numeric(layer_info$distance_outside_m))
  }

  lyr <- load_layer(
    layer_path,
    layer_type = if ("layer_type" %in% names(layer_info)) as.character(layer_info$layer_type) else "file",
    geometry_type = layer_geom,
    fields_to_extract = fields,
    bcdc_id = if ("bcdc_id" %in% names(layer_info)) as.character(layer_info$bcdc_id) else NA_character_,
    layer_name = if ("name" %in% names(layer_info)) as.character(layer_info$name) else layer_name,
    layer_identifier = if ("layer_identifier" %in% names(layer_info)) as.character(layer_info$layer_identifier) else NA_character_,
    search_dir = "data/input",
    cache_bcdc = TRUE
  )
  if (is.null(lyr)) {
    skipped_layers[[length(skipped_layers) + 1]] <- list(
      focus = layer_focus,
      layer_display = layer_display,
      geometry = layer_geom,
      reason = "Layer could not be loaded (missing local file or inaccessible BCDC source)"
    )
    next
  }

  direct_res <- NULL
  analysis_type <- NA_character_
  has_direct <- FALSE

  if (aoi_geom_type == "polygon") {
    if (layer_geom == "polygon") {
      direct_res <- polygon_overlap(aoi, lyr, fields)
      analysis_type <- "polygon_overlap"
      has_direct <- !is.null(direct_res)
    } else if (layer_geom == "line") {
      direct_res <- line_density_in_polygon(aoi, lyr, fields)
      analysis_type <- "line_density"
      has_direct <- !is.null(direct_res)
    } else if (layer_geom == "point") {
      direct_res <- points_in_polygon(aoi, lyr, fields)
      analysis_type <- "points_in_polygon"
      has_direct <- !is.null(direct_res)
    } else if (layer_geom == "raster") {
      direct_res <- raster_overlap(aoi, lyr)
      analysis_type <- "raster_overlap"
      has_direct <- !is.null(direct_res)
    }
  } else if (aoi_geom_type == "line") {
    if (layer_geom == "polygon") {
      direct_res <- line_polygon_overlap(aoi, lyr, fields)
      analysis_type <- "line_polygon_overlap"
      has_direct <- !is.null(direct_res)
    } else if (layer_geom == "line") {
      direct_res <- line_line_intersection(aoi, lyr, fields)
      analysis_type <- "line_intersection"
      has_direct <- !is.null(direct_res)
    } else if (layer_geom == "raster") {
      direct_res <- line_raster_overlap(aoi, lyr)
      analysis_type <- "line_raster_overlap"
      has_direct <- !is.null(direct_res)
    }
  }

  nearby_res <- NULL
  if (!is.na(outside_distance) && outside_distance > 0 && layer_geom != "raster") {
    nearby_res <- features_within_buffer(aoi, lyr, outside_distance, fields)
  }
  has_nearby <- !is.null(nearby_res)

  plot_vector <- NULL
  plot_raster <- NULL
  if (layer_geom == "raster") {
    plot_raster <- tryCatch(
      prepare_raster_plot_data(lyr, aoi, outside_distance),
      error = function(e) NULL
    )
  } else {
    context_pad <- if (is.na(outside_distance) || outside_distance <= 0) 500 else outside_distance

    if (has_direct) {
      plot_vector <- tryCatch(
        sf::st_filter(lyr, sf::st_union(aoi), .predicate = sf::st_intersects),
        error = function(e) NULL
      )
    }

    if ((is.null(plot_vector) || nrow(plot_vector) == 0) && has_nearby) {
      plot_vector <- tryCatch(
        sf::st_filter(lyr, sf::st_buffer(sf::st_union(aoi), dist = outside_distance), .predicate = sf::st_intersects),
        error = function(e) NULL
      )
    }

    if (!is.null(plot_vector) && nrow(plot_vector) > 0) {
      plot_vector <- tryCatch(
        crop_to_aoi_context(plot_vector, aoi, pad_dist = context_pad),
        error = function(e) plot_vector
      )
    } else {
      plot_vector <- NULL
    }
  }

  if (!has_direct && !has_nearby) {
    no_hit_layers[[length(no_hit_layers) + 1]] <- list(
      focus = layer_focus,
      layer_display = layer_display,
      geometry = layer_geom,
      outside_distance_m = outside_distance
    )
    rm(lyr, plot_vector, plot_raster)
    invisible(gc(verbose = FALSE))
    next
  }

  results[[layer_name]] <- format_result(layer_name, analysis_type, direct_res, has_direct)
  layer_outputs[[length(layer_outputs) + 1]] <- list(
    focus = layer_focus,
    layer_name = layer_name,
    layer_display = layer_display,
    layer_geom = layer_geom,
    spatial_need = spatial_need,
    analysis_type = analysis_type,
    direct = direct_res,
    has_direct = has_direct,
    map_geom = plot_vector,
    raster_plot_data = plot_raster,
    nearby = nearby_res,
    outside_distance = outside_distance
  )

  rm(lyr, plot_vector, plot_raster)
  invisible(gc(verbose = FALSE))
}

if (length(layer_outputs) == 0) {
  cat("*No configured layers intersected the AOI (or configured outside distances).*\n\n")
} else {
  focus_order <- unique(vapply(layer_outputs, function(x) x$focus, character(1)))

  for (focus in focus_order) {
    cat(sprintf("\n## Focus: %s\n\n", focus))
    focus_layers <- Filter(function(x) identical(x$focus, focus), layer_outputs)

    for (entry in focus_layers) {
      spatial_need_clean <- trimws(gsub("^\\*+|\\*+$", "", entry$spatial_need))
      cat(sprintf("### %s\n\n", entry$layer_display))
      cat(sprintf("**Geometry type:** %s\n\n", entry$layer_geom))
      cat(sprintf("**Requested spatial info:** %s\n\n", spatial_need_clean))

      if (isTRUE(params$interactive_maps)) {
        lmap <- build_layer_leaflet(
          aoi = aoi,
          aoi_geom_type = aoi_geom_type,
          layer_display = entry$layer_display,
          layer_geom = entry$layer_geom,
          plot_vector = entry$map_geom,
          plot_raster = entry$raster_plot_data
        )
        if (!is.null(lmap)) {
          cat("#### Analysis Map\n\n")
          render_leaflet_widget(lmap)
          cat("\n\n")
        }
      } else {
        layer_plot <- build_layer_plot(
          aoi = aoi,
          aoi_geom_type = aoi_geom_type,
          layer_display = entry$layer_display,
          layer_geom = entry$layer_geom,
          plot_vector = entry$map_geom,
          plot_raster = entry$raster_plot_data
        )
        if (!is.null(layer_plot)) {
          cat("#### Analysis Map\n\n")
          print(layer_plot)
          cat("\n\n")
        }
      }

      if (entry$has_direct) {
        if (entry$analysis_type == "polygon_overlap") {
          cat("#### Overlap details\n\n")
          polygon_totals <- attr(entry$direct, "coverage_summary", exact = TRUE)
          if (is.null(polygon_totals)) {
            polygon_totals <- data.frame(
              total_overlap_area_ha = sum(entry$direct$overlap_area_ha, na.rm = TRUE),
              total_overlap_pct = if (!is.na(aoi_area_ha) && aoi_area_ha > 0) {
                (sum(entry$direct$overlap_area_ha, na.rm = TRUE) / aoi_area_ha) * 100
              } else {
                NA_real_
              }
            )
          }
          { render_table(polygon_totals, caption = paste(entry$layer_display, "- Total Overlap")); invisible(NULL) }
          cat("\n\n")
          { render_grouped_tables(entry$direct, caption_prefix = paste(entry$layer_display, "- Overlap")); invisible(NULL) }
          cat("\n\n")
        } else if (entry$analysis_type == "line_density") {
          cat("#### Line density and type\n\n")
          { render_table(entry$direct$summary, caption = paste(entry$layer_display, "- Total Length and Density")); invisible(NULL) }
          cat("\n\n")
          { render_grouped_tables(entry$direct$details, caption_prefix = paste(entry$layer_display, "- Features")); invisible(NULL) }
          cat("\n\n")
        } else if (entry$analysis_type == "points_in_polygon") {
          cat(sprintf("#### Points in AOI: %d\n\n", entry$direct$count))
          points_totals <- data.frame(total_point_count = entry$direct$count)
          { render_table(points_totals, caption = paste(entry$layer_display, "- Total Points")); invisible(NULL) }
          cat("\n\n")
          { render_grouped_tables(entry$direct$details, caption_prefix = paste(entry$layer_display, "- Points")); invisible(NULL) }
          cat("\n\n")
        } else if (entry$analysis_type == "raster_overlap") {
          cat("#### Raster value percentages in AOI\n\n")
          raster_totals <- data.frame(
            raster_classes = nrow(entry$direct),
            total_cells = if ("count" %in% names(entry$direct)) sum(entry$direct$count, na.rm = TRUE) else NA_real_,
            total_pct_of_aoi = if ("pct_of_aoi" %in% names(entry$direct)) sum(entry$direct$pct_of_aoi, na.rm = TRUE) else NA_real_
          )
          { render_table(raster_totals, caption = paste(entry$layer_display, "- Total Raster Coverage")); invisible(NULL) }
          cat("\n\n")
          { render_table(entry$direct, caption = paste(entry$layer_display, "- Raster Overlap")); invisible(NULL) }
          cat("\n\n")
        } else if (entry$analysis_type == "line_polygon_overlap") {
          cat("#### Line overlap length\n\n")
          line_polygon_totals <- attr(entry$direct, "coverage_summary", exact = TRUE)
          if (is.null(line_polygon_totals)) {
            overlap_total_km <- sum(entry$direct$overlap_length_km, na.rm = TRUE)
            line_polygon_totals <- data.frame(
              total_overlap_length_km = overlap_total_km,
              total_pct_of_aoi_line = if (!is.na(aoi_length_km) && aoi_length_km > 0) (overlap_total_km / aoi_length_km) * 100 else NA_real_
            )
          }
          { render_table(line_polygon_totals, caption = paste(entry$layer_display, "- Total Overlap Length")); invisible(NULL) }
          cat("\n\n")
          { render_grouped_tables(entry$direct, caption_prefix = paste(entry$layer_display, "- Overlap Length")); invisible(NULL) }
          cat("\n\n")
        } else if (entry$analysis_type == "line_intersection") {
          cat("#### Intersecting linear features\n\n")
          intersection_totals <- data.frame(total_intersection_count = sum(entry$direct$intersection_count, na.rm = TRUE))
          { render_table(intersection_totals, caption = paste(entry$layer_display, "- Total Intersections")); invisible(NULL) }
          cat("\n\n")
          { render_grouped_tables(entry$direct, caption_prefix = paste(entry$layer_display, "- Intersections")); invisible(NULL) }
        } else if (entry$analysis_type == "line_raster_overlap") {
          cat("#### Raster overlap length along line\n\n")
          line_raster_totals <- data.frame(
            total_overlap_length_km = sum(entry$direct$overlap_length_km, na.rm = TRUE),
            total_pct_of_line = if ("pct_of_line" %in% names(entry$direct)) sum(entry$direct$pct_of_line, na.rm = TRUE) else NA_real_
          )
          { render_table(line_raster_totals, caption = paste(entry$layer_display, "- Total Raster Overlap")); invisible(NULL) }
          cat("\n\n")
          { render_table(entry$direct, caption = paste(entry$layer_display, "- Raster Along Line")); invisible(NULL) }
          cat("\n\n")
        }
      }

      if (!is.null(entry$nearby)) {
        cat(sprintf("#### Features within %.0f m outside AOI\n\n", entry$outside_distance))
        if ("feature_count" %in% names(entry$nearby)) {
          total_nearby <- sum(entry$nearby$feature_count, na.rm = TRUE)
          weighted_mean_dist <- if (total_nearby > 0 && "mean_distance_m" %in% names(entry$nearby)) {
            sum(entry$nearby$mean_distance_m * entry$nearby$feature_count, na.rm = TRUE) / total_nearby
          } else {
            NA_real_
          }
          nearby_totals <- data.frame(
            total_nearby_features = total_nearby,
            min_distance_m = if ("min_distance_m" %in% names(entry$nearby)) min(entry$nearby$min_distance_m, na.rm = TRUE) else NA_real_,
            mean_distance_m = weighted_mean_dist,
            max_distance_m = if ("max_distance_m" %in% names(entry$nearby)) max(entry$nearby$max_distance_m, na.rm = TRUE) else NA_real_
          )
        } else {
          nearby_totals <- data.frame(
            total_nearby_features = nrow(entry$nearby),
            min_distance_m = if ("distance_m" %in% names(entry$nearby)) min(entry$nearby$distance_m, na.rm = TRUE) else NA_real_,
            mean_distance_m = if ("distance_m" %in% names(entry$nearby)) mean(entry$nearby$distance_m, na.rm = TRUE) else NA_real_,
            max_distance_m = if ("distance_m" %in% names(entry$nearby)) max(entry$nearby$distance_m, na.rm = TRUE) else NA_real_
          )
        }
        { render_table(nearby_totals, caption = paste(entry$layer_display, "- Nearby Feature Totals")); invisible(NULL) }
        cat("\n\n")
        { render_grouped_tables(entry$nearby, caption_prefix = paste(entry$layer_display, "- Nearby Features")); invisible(NULL) }
        cat("\n\n")
      }
    }
  }
}

```

# Layer Exclusions and Load Issues

```{r layer-exclusions, results='asis'}
if (length(no_hit_layers) > 0) {
  cat("## Layers With No AOI Match\n\n")
  no_hit_df <- data.frame(
    Focus = vapply(no_hit_layers, function(x) x$focus, character(1)),
    Layer = vapply(no_hit_layers, function(x) x$layer_display, character(1)),
    Geometry = vapply(no_hit_layers, function(x) x$geometry, character(1)),
    outside_distance_m = vapply(no_hit_layers, function(x) ifelse(is.na(x$outside_distance_m), NA, x$outside_distance_m), numeric(1))
  )
  { render_table(no_hit_df, caption = "Configured layers with no direct overlap and no nearby features"); invisible(NULL) }
  cat("\n\n")
}

if (length(skipped_layers) > 0) {
  cat("## Skipped Layers (Load Issues)\n\n")
  skipped_df <- data.frame(
    Focus = vapply(skipped_layers, function(x) x$focus, character(1)),
    Layer = vapply(skipped_layers, function(x) x$layer_display, character(1)),
    Geometry = vapply(skipped_layers, function(x) x$geometry, character(1)),
    Reason = vapply(skipped_layers, function(x) x$reason, character(1))
  )
  { render_table(skipped_df, caption = "Configured layers skipped because source data could not be loaded"); invisible(NULL) }
  cat("\n\n")
}
```

# Summary

```{r summary-table, results='asis'}
if (length(layer_outputs) == 0) {
  cat("*No layer hits to summarise.*\n\n")
} else {
  summary_df <- data.frame(
    Focus = vapply(layer_outputs, function(x) x$focus, character(1)),
    Layer = vapply(layer_outputs, function(x) x$layer_display, character(1)),
    Geometry = vapply(layer_outputs, function(x) x$layer_geom, character(1)),
    Spatial_Request = vapply(layer_outputs, function(x) x$spatial_need, character(1)),
    Direct_Hit = vapply(layer_outputs, function(x) ifelse(x$has_direct, "Yes", "No"), character(1)),
    Outside_Distance_m = vapply(layer_outputs, function(x) ifelse(is.na(x$outside_distance), NA, x$outside_distance), numeric(1)),
    Nearby_Hit = vapply(layer_outputs, function(x) ifelse(is.null(x$nearby), "No", "Yes"), character(1))
  )
  { render_table(summary_df, caption = "Detected layer results by focus"); invisible(NULL) }
}
```

---

*Report generated by the BC Gov Referral Spatial Analysis tool.*
*See README.md for instructions on updating reference layers and running this report.*
